Empirical Bayes analysis for detection of gene heterosis in RNAseq data

### 2014-12-16 ebayes_rnaseq_heterosis.tex
start of introduction

\begin{keywords}
Empirical Bayes; Heterosis; Hierarchical Model; Negative binomial; RNAseq.
\end{keywords}

@@

\section{Heterosis}
\label{s:heterosis}

Well before heterosis was scientifically described by \cite{darwin1876effects}, humans had been using heterosis for various practical purposes. Within the last century, heterosis has been used to improve many species for food, feed, and fuel industries. A few relatively recent examples include rice \cite{yu1997importance}, alfalfa \cite{riday2002heterosis}, tomatoes \cite{krieger2010flowering}, and fish \cite{wohlfarth1993heterosis}. Despite the intensive study and successful utilization of heterosis, the basic genomic mechanisms responsible for heterosis remain unclear \cite{lippman2007heterosis}.

\cite{shull1952beginnings} coined the term heterosis and defined it as 
\begin{quote}
the interpretation of increased vigor, size, fruitfulness, speed of development, resistance to disease and to insect pests, or to climatic rigors of any kind, manifested by crossbred organisms as compared with corresponding inbreds, as the specific results of unlikeness in the constitutions of the uniting parental gametes. 
\end{quote}

For example, the maize hybrid produced from a cross of parental inbred lines B73 and Mo17 is larger in size, has lower maturation time, and has higher grain yield than both of its parents \cite{hallauer2010quantitative}. Data from \citep{paschold2012complementation}

\citep{ji2014estimation}

@@

The {\tt ShrinkBayes} package implements fully Bayesian inference through integrated nested Laplace approximation (INLA) \citep{rue2009approximate,van2014shrinkbayes}. A limitation with INLA is that it only provides approximations to marginal posteriors rather than the full posterior. Thus we cannot directly estimate nonlinear quantities such as $P(\mu_{g3} < \min(\mu_{g1},\mu_{g2})$. To circumvent this limitation, we reparameterized the independence model in Section \ref{s:model} following \cite{ji2014estimation}. Let $\phi_g = (\mu_{g1}+\mu_{g2})/2$ be the parental log-mean, $\alpha_g = (\mu_{g2}-\mu_{g1})/2$ be the half parental difference, and $\delta_g = \mu_{g3} - \phi_g$ be the hybrid effect. We implemented this model using {\tt ShrinkBayes} and constructed a statistic based on posterior median of these parameters. Specifically, the statistic is 0 if $|\hat{\delta}_f| \le |\hat{\alpha}_f|$, $P(\delta_f + \alpha_f > 0 |y)$ if $\hat{\delta}_f > -\hat{\alpha}_f \ge 0$, $P(\delta_f - \alpha_f > 0 |y)$ if $\hat{\delta}_f > \hat{\alpha}_f \ge 0$, $P(\delta_f - \alpha_f < 0 |y)$ if $\hat{\delta}_f < \hat{\alpha}_f \le 0$, and $P(\delta_f + \alpha_f < 0 |y)$ if $\hat{\delta}_f < -\hat{\alpha}_f \le 0$.

@@

Figure \ref{f:auc} provides the area under the ROC below a false positive rate of 0.1 across the 10 simulations for each of the 3 different sample sizes. 
\begin{figure}
\centerline{\includegraphics[width=\textwidth]{auc-facet-TRUE}}
\caption{Area under the ROC curves below a false positive rate of 0.1 for 3 different sample sizes for the approaches using {\tt edgeR}, {\tt baySeq},  and {\tt ShrinkBayes} described in Section \ref{s:alternative} and the method using {\tt RStan} assuming independence and with a covariance described in Section \ref{s:gene_heterosis}. Each line is a different simulation while the blue box indicates mean AUC and its standard error across the simulations.}
\label{f:auc}
\end{figure}
The pattern appears to hold with the {\tt RStan} and {\tt ShrinkBayes} approaches outperforming the other two. While there does not appear to be a significant difference between the independence model using RStan or ShrinkBayes, there does appear to be a significant decreased performance, at least for sample sizes, when using the model where a covariance is estimated.

@@

The authors thank Andrew Lithio for help implementing our model in {\tt ShrinkBayes}.

>>>>

\begin{keywords}
Empirical Bayes; Gene expression; Heterosis; Hierarchical model; Negative binomial; RNAseq.
\end{keywords}

@@

% Need to udpate as this is directly from Tieming's paper
Heterosis, or hybrid vigor, refers to the enhances phenotype of hybrid progeny relative to their inbred parents. Taking maize as an example, the offspring from crossing the inbred lines B73 and Mo17 are taller, mature faster, and produce greater yields than their parent lines \citep{hallauer1981quantitative, hallauer2010quantitative}. Since heterosis was scientifically documented by \cite{darwin1876effects}, it has been successfully manipulated to improve many species for food, feed, and fuel industries, such as rice \citep{yu1997importance}, alfalfa \citep{riday2002heterosis}, tomatoes \citep{krieger2010flowering}, and fish \citep{wohlfarth1993heterosis}. Despite the intensive study and successful utilization of heterosis, the basic genomic mechanisms remain unclear \citep{coors1999genetics, lippman2007heterosis}. Researchers speculate that gene-expression heterosis could be among the mechanisms responsible for the phenotypic heterosis \citep{swanson2006all, springer2007allelic}.

Recently \citep{ji2014estimation} provided an approach to assess heterosis using micro-array data using a hierarchical model for the ...

%\section{Heterosis}
%\label{s:heterosis}

% Well before heterosis was scientifically described by , humans had been using heterosis for various practical purposes. Within the last century, heterosis has been used to improve many species for food, feed, and fuel industries. A few relatively recent examples include rice , alfalfa , tomatoes . Despite the intensive study and successful utilization of heterosis, the basic genomic mechanisms responsible for heterosis remain unclear .
% 
% \cite{shull1952beginnings} coined the term heterosis and defined it as 
% \begin{quote}
% the interpretation of increased vigor, size, fruitfulness, speed of development, resistance to disease and to insect pests, or to climatic rigors of any kind, manifested by crossbred organisms as compared with corresponding inbreds, as the specific results of unlikeness in the constitutions of the uniting parental gametes. 
% \end{quote}
% 
% For example, the maize hybrid produced from a cross of parental inbred lines B73 and Mo17 is larger in size, has lower maturation time, and has higher grain yield than both of its parents \cite{hallauer2010quantitative}. Data from \citep{paschold2012complementation}
% 
% \citep{ji2014estimation}

@@

The {\tt ShrinkBayes} package implements fully Bayesian inference through integrated nested Laplace approximation (INLA) \citep{rue2009approximate,van2014shrinkbayes}. A limitation with INLA is that it only provides approximations to marginal posteriors for parameters or linear combinations of parameters rather than an approximation to the full posterior. Thus, we cannot directly estimate nonlinear quantities such as $P(\mu_{g3} < \min(\mu_{g1},\mu_{g2}))$. To circumvent this limitation, we reparameterized the independence model in Section \ref{s:model} following \cite{ji2014estimation}. Let $\phi_g = (\mu_{g1}+\mu_{g2})/2$ be the parental log-mean, $\alpha_g = (\mu_{g2}-\mu_{g1})/2$ be the half parental difference, and $\delta_g = \mu_{g3} - \phi_g$ be the hybrid effect. We estimated this model using {\tt ShrinkBayes} and constructed a statistic based on the posterior mean of these parameters. Specifically, the statistic is 0 if $|\hat{\delta}_f| \le |\hat{\alpha}_f|$, $P(\delta_f + \alpha_f > 0 |y)$ if $\hat{\delta}_f > -\hat{\alpha}_f \ge 0$, $P(\delta_f - \alpha_f > 0 |y)$ if $\hat{\delta}_f > \hat{\alpha}_f \ge 0$, $P(\delta_f - \alpha_f < 0 |y)$ if $\hat{\delta}_f < \hat{\alpha}_f \le 0$, and $P(\delta_f + \alpha_f < 0 |y)$ if $\hat{\delta}_f < -\hat{\alpha}_f \le 0$.

@@

Figure \ref{f:auc} provides the area under the ROC below a false positive rate of 0.1 across the 10 simulations for each of the 3 different sample sizes. 
\begin{figure}
\centerline{\includegraphics[width=\textwidth]{auc-facet-TRUE}}
\caption{Area under the ROC curves below a false positive rate of 0.1 for 3 different sample sizes for the approaches using {\tt edgeR}, {\tt baySeq},  and {\tt ShrinkBayes} described in Section \ref{s:alternative} and the method using {\tt RStan} assuming independence and with a covariance described in Section \ref{s:gene_heterosis}. Each line is a different simulation while the blue box indicates mean AUC and its standard error across the simulations.}
\label{f:auc}
\end{figure}
The pattern appears to hold with the {\tt RStan} and {\tt ShrinkBayes} approaches outperforming the other two. While there does not appear to be a significant difference between the independence model using RStan or ShrinkBayes, there does appear to be a significant decreased performance, at least for smaller sample sizes, when using the model where a covariance is estimated.

@@

The authors thank Andrew Lithio for help implementing our model in {\tt ShrinkBayes}. Research reported in this publication was supported by National Institute of General Medical Sciences of the National Institutes of Health under award number R01GM109458. The content is solely the responsibility of the authors and does not necessarily represent the official views of the National Institutes of Health.
